prolfqua - R-package for Analysing Proteomics Label Free Quantification Experiments

### 2022-01-24 inst/MyArticle/MergePoster.Rmd
itration intro

When analyzing parallel designs using a single explanatory variable, we can use all packages for the statistical data analysis of the data; however, if we want to model factorial designs or repeated measurements, we can use only some of them. Table \ref@(tab:tableOverview) gives an overview of the models and features supported by these packages.

@@

We were inspired by the excellent R package caret, which enables to call various machine learning methods, making selecting the best ML algorithm for your problem easy. We wanted to have a similar tool for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, all of them fit some type of linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, some of them apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular design, with R linear models as a core, where features such as p-value aggregation (ROPECA) or variance shrinkage could be easily enabled if required.

Models in _prolfqua_ are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with _prolfqua_ using base R or other programming languages by reading the analysis script and without looking at the package code. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of MSstats, we provide the functionality to suggest a model formula from the sample annotation.

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. However, if there are not observation in one of the groups, then contrasts involving this group can not be estimated. To adres it, for the group without observations, we use a point prior equal to the average of $10\%$ smallest group mean intensities in the dataset and infer the variance from the variance in other groups. A way to adress this is to annotate the produced models with information about missigness which is done by for instance MSStats. How missigness can modelled using the Hurdle model is described here, however to our knowledge only the R package _proDA_ models missingness directly.

Furthermore, the functionality of _prolfqua_ also includes methods specific to proteomics data. For example, we implemented strategies to infer protein intensities from peptide intensities: top N, Tukey's median polish, robust linear models. Furthermore, peptide or protein intensities can be scaled using the robust scaling method or transformed using _quantile_ normalization or _vsn_ from Bioconductor to remove systematic differences among samples and heteroscedasticity. In this respect, _prolfqua_ can be compared with other R packages such _DEP_ [@DEP2018] or POMA [@POMA2021] which support the entire LFQ analysis pipeline.

>>>>

TODO link to table \@ref(tab:tableOverview)

When analyzing parallel group designs using a single explanatory variable, we can use all packages for the statistical data analysis of the data; however, if we want to model factorial designs or repeated measurements, we can use only some of them. Table \ref@(tab:tableOverview) gives an overview of the models and features supported by these packages.

@@

We were inspired by the R package caret [TODO cite], which enables to call various machine learning methods, making selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, all of them fit some type of linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, some of them apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular design (TODO - clarity), with R linear models as a core, where features such as p-value aggregation (ROPECA) or variance shrinkage could be easily enabled if required.

Models in _prolfqua_ are specified using R’s linear model and mixed model formula interface. Therefore, knowledge of the R regression model infrastructure (TODO cite) is an advantage when using our package. Furthermore, this glass box approach should make it easy to reimplement an analysis performed with _prolfqua_ using base R or other programming languages by reading the analysis script and without looking at the package code. However, acknowledging the R formula interface’s complexity to non-statisticians and the popularity of _MSstats_, we provide the functionality to suggest a model formula from the sample annotation.

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. However, if there are not observation in one of the groups, then contrasts involving this group can not be estimated. To address this, for the group without observations, we use a point prior equal to the average of $1\%$ smallest group mean intensities in the dataset and infer the variance from the variance in other groups. (TODO use different start of sentence) A way to address this is to annotate the produced models with information about missingness which is done by for instance _MSStats_. How missingness can modelled using the Hurdle model is described here (TODO citations), however to our knowledge only the R package _proDA_ models missingness directly.

Furthermore, the functionality of _prolfqua_ also includes methods specific to proteomics data. For example, we implemented strategies to estimate protein intensities from peptide intensities: top N, Tukey's median polish, robust linear models (cite?). Furthermore, peptide or protein intensities can be scaled using the robust scaling method or transformed using _quantile_ normalization or _vsn_ from Bioconductor to remove systematic differences among samples and heteroscedasticity. In this respect, _prolfqua_ can be compared with other R packages such _DEP_ [@DEP2018] or POMA [@POMA2021] which support the entire LFQ analysis pipeline.
